---
title: "Practice Sheet Day 4"
subtitle: "SARA Summer School"
author: "Dr. Ajay Koli"
date: today
format: html
toc: true
number-sections: true
---

## Variables

We’ll simulate data to demonstrate the relationship between independent and dependent variables.

### Example: Hours of Study and Test Scores

```{r}
# Simulate data
set.seed(123)
hours_of_study <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10) # Independent Variable
test_scores <- 50 + 5 * hours_of_study + rnorm(10, mean = 0, sd = 5) # Dependent Variable

# Combine into a data frame
data <- data.frame(Hours = hours_of_study, Scores = test_scores)

# Visualize the relationship
library(ggplot2)

ggplot(data, aes(x = Hours, y = Scores)) +
  geom_point(color = "blue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Effect of Hours of Study on Test Scores",
    x = "Hours of Study (Independent Variable)",
    y = "Test Scores (Dependent Variable)"
  ) +
  theme_minimal()

```

Interpretation of the Graph

- X-axis (Independent Variable): Hours of study.
    
- Y-axis (Dependent Variable): Test scores.
    
- As hours of study increase, test scores generally increase, showing a positive relationship.

### Inverse Relationship Example

An inverse relationship occurs when one variable increases while the other decreases. For instance, the more time spent on social media, the lower the grades a student might achieve.

```{r}
# Simulate data
set.seed(42)
social_media_hours <- seq(1, 10, by = 1)  # Independent Variable
grades <- 100 - 5 * social_media_hours + rnorm(10, mean = 0, sd = 2)  # Dependent Variable

# Combine into a data frame
inverse_data <- data.frame(SocialMediaHours = social_media_hours, Grades = grades)

# Visualize the inverse relationship
library(ggplot2)

ggplot(inverse_data, aes(x = SocialMediaHours, y = Grades)) +
  geom_point(color = "red", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title = "Inverse Relationship: Social Media Hours vs. Grades",
    x = "Social Media Hours (Independent Variable)",
    y = "Grades (Dependent Variable)"
  ) +
  theme_minimal()

```

### No-Relation Example

A no-relation scenario means that changes in one variable do not systematically affect the other. For instance, the number of books read by a person and their shoe size typically have no relationship.

```{r}
# Simulate data
set.seed(123)
books_read <- seq(1, 10, by = 1)  # Independent Variable
shoe_size <- rnorm(10, mean = 9, sd = 1)  # Random data with no relation to books_read

# Combine into a data frame
no_relation_data <- data.frame(BooksRead = books_read, ShoeSize = shoe_size)

# Visualize the no-relation scenario
ggplot(no_relation_data, aes(x = BooksRead, y = ShoeSize)) +
  geom_point(color = "green", size = 3) +
  labs(
    title = "No Relationship: Books Read vs. Shoe Size",
    x = "Books Read (Independent Variable)",
    y = "Shoe Size (Dependent Variable)"
  ) +
  theme_minimal()

```

## NHST

### One-Sample t-Test

**Scenario:** A teacher claims that the average test score in a class is 70. You want to test if the actual average is different from 70.

- $H_0$: The mean test score is 70 ($\mu$ = 70).

- $H_a$: The mean test score is not 70 ($\mu \neq$  70).

```{r}
# Simulated data: Test scores of students
set.seed(123)
test_scores <- rnorm(30, mean = 72, sd = 5)

# Perform a one-sample t-test
t_test_result <- t.test(test_scores, mu = 70)

# Print the result
print(t_test_result)

```

Interpretation:

- If the p-value is less than 0.05, reject $H_0$. This means the average test score is significantly different from 70.

### Two-Sample t-Test

Scenario: Compare the average test scores of two groups: one that received tutoring and one that did not.

- $H_0$: The mean test scores the two groups are equal ($\mu_1 = \mu_2$).

- $H_a$: The mean test scores of the two groups are not equal ($\mu_1 \neq  \mu_2$).

```{r}
# Simulated data
set.seed(456)
group_A <- rnorm(30, mean = 75, sd = 5)  # Tutored group
group_B <- rnorm(30, mean = 70, sd = 5)  # Non-tutored group

# Perform a two-sample t-test
two_sample_test <- t.test(group_A, group_B, alternative = "two.sided")

# Print the result
print(two_sample_test)

```

Interpretation:

- If the p-value is less than 0.05, reject $H_0$. This indicates a significant difference between the two groups.

### Chi-Square Test

Scenario: Test if there is an association between gender and preference for two types of beverages.

- $H_0$: Gender and beverage preference are independent.

- $H-a$: Gender and beverage preference are not independent.

```{r}
# Simulated data
gender <- c("Male", "Male", "Female", "Female", "Male", "Female", "Male", "Female")
beverage <- c("Tea", "Coffee", "Tea", "Coffee", "Coffee", "Tea", "Coffee", "Tea")

# Create a contingency table
table_data <- table(gender, beverage)

# Perform a chi-square test
chi_sq_test <- chisq.test(table_data)

# Print the result
print(chi_sq_test)

```

Interpretation:

- If the p-value is less than 0.05, reject $H_0$. This suggests that gender and beverage preference are associated.

### ANOVA (Analysis of Variance)

Scenario: Test if three different fertilizers lead to different crop yields.

- $H_0$: The mean yields are the same for all fertilizers ($\mu_1 = \mu_2 = \mu_3$).

- $H_a$: At least one fertilizer leads to a different mean yield

```{r}
# Simulated data
set.seed(789)
fertilizer <- factor(rep(c("Fert1", "Fert2", "Fert3"), each = 10))
yield <- c(rnorm(10, mean = 50, sd = 5), rnorm(10, mean = 55, sd = 5), rnorm(10, mean = 60, sd = 5))

# Perform ANOVA
anova_result <- aov(yield ~ fertilizer)
summary(anova_result)

```

Interpretation:

- If the p-value is less than 0.05, reject $H_0$. This means at least one fertilizer has a significantly different effect.

### Correlation Test

Scenario: Test if there is a significant correlation between hours studied and test scores.

- $H_0$: There is no correlation between hours studied and test scores ($r = 0$).
    
- $H_a$: There is a correlation between hours studied and test scores ($r \neq 0$).

```{r}
# Simulated data
set.seed(101)
hours_studied <- rnorm(50, mean = 5, sd = 2)
test_scores <- hours_studied * 10 + rnorm(50, mean = 50, sd = 10)

# Perform correlation test
cor_test <- cor.test(hours_studied, test_scores)

# Print the result
print(cor_test)

```

Interpretation:

- If the p-value is less than 0.05, reject $H_0$. This indicates a significant correlation.

## Correlation

### Positive Correlation

When one variable increases, the other variable also increases.

Example: Height and weight of individuals.

```{r}
# Simulated data
set.seed(123)
height <- rnorm(50, mean = 160, sd = 10)  # Heights in cm
weight <- height * 0.5 + rnorm(50, mean = 50, sd = 5)  # Weights in kg

# Calculate correlation
correlation_positive <- cor(height, weight)

# Visualization
library(ggplot2)
ggplot(data.frame(height, weight), aes(x = height, y = weight)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  ggtitle(paste("Positive Correlation: r =", round(correlation_positive, 2))) +
  theme_minimal()

```

### Negative Correlation

When one variable increases, the other variable decreases.

Example: Hours spent on social media vs. academic performance.

```{r}
# Simulated data
set.seed(456)
social_media_hours <- rnorm(50, mean = 5, sd = 1.5)
academic_performance <- 100 - (social_media_hours * 10) + rnorm(50, mean = 0, sd = 5)

# Calculate correlation
correlation_negative <- cor(social_media_hours, academic_performance)

# Visualization
ggplot(data.frame(social_media_hours, academic_performance), aes(x = social_media_hours, y = academic_performance)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  ggtitle(paste("Negative Correlation: r =", round(correlation_negative, 2))) +
  theme_minimal()

```

### Zero Correlation

No relationship exists between the variables.

Example: Shoe size and IQ.

```{r}
# Simulated data
set.seed(789)
shoe_size <- rnorm(50, mean = 8, sd = 1.5)
iq_scores <- rnorm(50, mean = 100, sd = 15)

# Calculate correlation
correlation_zero <- cor(shoe_size, iq_scores)

# Visualization
ggplot(data.frame(shoe_size, iq_scores), aes(x = shoe_size, y = iq_scores)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  ggtitle(paste("Zero Correlation: r =", round(correlation_zero, 2))) +
  theme_minimal()

```

