---
title: "Data Visualisation with R"
subtitle: "Session 1 · Getting Started — R, RStudio & Your First Plot"
author: "SARA Institute of Data Science"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    incremental: false
    toc: false
    footer: "SARA Institute of Data Science · R for Researchers"
    logo: ""
    scrollable: true
execute:
  echo: true
  warning: false
  message: false
---

## Welcome! {.center}

::: {.incremental}
- This is a **hands-on** workshop. Please have RStudio open in front of you.
- **No prior coding experience is assumed.** We start from zero.
- Three sessions, ~90 minutes each:
  1. Getting Started — R, RStudio, your first plot
  2. Grammar of Graphics — aesthetics, geoms, facets
  3. Polishing & Publishing — labels, themes, saving your work
- Today's motto: **you will make a real plot in the first 30 minutes.**
:::

## Why data visualisation?

- A good chart answers a question **faster than a table of numbers ever could**
- As researchers, we visualise to:
  - **explore** data and spot patterns/errors
  - **explain** findings to others (papers, viva, funders, community)
- R + `ggplot2` is one of the most widely used tools for this in research, government, and industry worldwide

. . .

> Goal for today: understand *why* a plot looks the way it does, not just copy-paste code.

## Session 1 roadmap {.smaller}

| Time | Topic |
|---|---|
| 0–10 min | What is R? What is RStudio? |
| 10–25 min | Panes tour + first R commands |
| 25–40 min | Packages, loading data, exploring `penguins` |
| 40–55 min | The Grammar of Graphics + first scatter plot |
| 55–75 min | Aesthetics: colour, and bar charts |
| 75–90 min | Your Turn practice + wrap-up |

# What is R? What is RStudio?

## R vs RStudio — an analogy

::: {.columns}
::: {.column width="50%"}
**R**

- The *engine* — the actual programming language
- Free and open-source
- Does all the actual computing
:::

::: {.column width="50%"}
**RStudio**

- The *car* built around the engine
- A friendly interface (IDE) to write and run R code
- Also free and open-source
:::
:::

. . .

You always need **R installed first**, then **RStudio** on top of it. You will never open "R" directly — you always work through RStudio.

## A tour of RStudio {.smaller}

When you open RStudio you will usually see **four panes**:

1. **Source (top-left)** — where you write and save your code (a script, `.R` or `.qmd` file)
2. **Console (bottom-left)** — where code actually runs, one line at a time
3. **Environment / History (top-right)** — shows objects you've created (data, variables)
4. **Files / Plots / Packages / Help (bottom-right)** — file browser, and **this is where your plots will appear!**

. . .

**Golden rule:** always write your code in the **Script**, not directly in the Console. The script is saved; the console is not.

## Your very first R commands

Type these into the **Console** and press Enter after each line:

```{r}
2 + 2
```

```{r}
"Hello, SARA!"
```

```{r}
10 / 3
```

. . .

R is a calculator, a text-handler, and — very soon — a plotting machine, all at once.

## Creating objects with `<-`

In R, we **store** things using the assignment arrow `<-` (shortcut: `Alt` + `-`)

```{r}
my_age <- 25
my_age
```

```{r}
workshop_name <- "R for Researchers"
workshop_name
```

. . .

Think of `<-` as **"put this value into this box, and name the box."** You can reuse the box (object) any time by typing its name.

## A note on scripts vs `.qmd` files

- A plain **R script** (`.R`) runs code top to bottom
- A **Quarto document** (`.qmd`) — like the one this workshop is built in! — mixes **writing** and **code** together, and can produce slides, reports, PDFs, or websites
- Code lives inside "chunks":

````
```{{r}}
2 + 2
```
````

. . .

For this workshop, a simple **R Script** (`File → New File → R Script`) is all you need. We'll use `.qmd` later if time allows.

# Packages, and Meeting Our Data

## What is a "package"?

- Base R can do a lot, but **packages** add extra tools written by the R community
- Install a package **once** per computer:

```{r}
#| eval: false
install.packages("tidyverse")
install.packages("palmerpenguins")
```

- Then **load** it every time you start a new R session:

```{r}
library(tidyverse)
library(palmerpenguins)
```

. . .

`install.packages()` = *buying the toolbox*. `library()` = *opening the toolbox on your desk today.*

## Meet the tidyverse

`tidyverse` is not one package — it's a **collection** of packages that work well together, including:

- `ggplot2` — for **data visualisation** (our focus!)
- `dplyr` — for filtering/summarising data
- `readr` — for reading in data files (CSV etc.)

. . .

Loading `tidyverse` loads all of these together, so we only need one `library()` line.

## Meet the data: `palmerpenguins`

- A free, beginner-friendly dataset of **344 penguins** from three species, collected near Palmer Station, Antarctica
- Comes bundled inside the `palmerpenguins` package — no file to download!
- Variables include:

| Variable | Meaning |
|---|---|
| `species` | Adelie, Chinstrap, or Gentoo |
| `island` | Biscoe, Dream, or Torgersen |
| `bill_length_mm` / `bill_depth_mm` | beak size |
| `flipper_length_mm` | flipper size |
| `body_mass_g` | weight in grams |
| `sex`, `year` | sex and year recorded |

## Loading and looking at the data

```{r}
data(penguins)
head(penguins)
```

. . .

```{r}
glimpse(penguins)
```

## Quick exploration commands

```{r}
dim(penguins)      # rows, columns
```

```{r}
summary(penguins$body_mass_g)   # quick stats on one column
```

. . .

**Note the `$` sign** — `penguins$body_mass_g` means *"go into the `penguins` table and pull out the `body_mass_g` column."*

## Your Turn 1 (5 min) {.smaller}

1. Load the `penguins` data (if not already loaded)
2. Use `str()` on `penguins` — how is it different from `glimpse()`?
3. Find the summary of `flipper_length_mm` using `summary()`

. . .

**Answer**

```{r}
str(penguins)
```

```{r}
summary(penguins$flipper_length_mm)
```

`str()` and `glimpse()` show similar info; `glimpse()` (from `dplyr`) is usually easier to read because it prints one variable per line.

# The Grammar of Graphics

## The big idea behind `ggplot2`

"gg" stands for **Grammar of Graphics** — every plot is built from the same simple *layers*:

::: {.incremental}
1. **Data** — which data frame?
2. **Aesthetics (`aes`)** — which columns map to x, y, colour, size...?
3. **Geometry (`geom`)** — what *type* of plot? points, bars, lines...
4. (later: facets, themes, labels, scales)
:::

. . .

Once you know this grammar, you can build **almost any chart** by combining the same small set of building blocks.

## Your first plot — a scatter plot

```{r}
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

## Breaking that down, line by line

```r
ggplot(data = penguins,                              # 1. which data?
       aes(x = flipper_length_mm, y = body_mass_g)) +# 2. what maps to x/y?
  geom_point()                                        # 3. draw it as points
```

- `ggplot()` starts the plot and sets up the "canvas"
- `aes()` = **aesthetic mapping** — connects data columns to visual properties
- `+` **adds a layer** — geoms are always added with `+`, never a comma
- `geom_point()` says: "draw one point per row of data"

. . .

**Common beginner mistake:** putting `+` at the *start* of a new line instead of the *end* of the previous line. R needs the `+` at the end to know more is coming!

## Your Turn 2 (7 min) {.smaller}

Make a scatter plot of `bill_length_mm` (x-axis) against `bill_depth_mm` (y-axis) using the `penguins` data.

*Hint: copy the previous code and change the column names.*

. . .

**Answer**

```{r}
ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point()
```

Do you notice the points seem to form **three separate clusters**? We'll find out why on the next slide!

## Adding colour by species

```{r}
ggplot(data = penguins,
       aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point()
```

. . .

Mystery solved — the three "clusters" are actually the **three penguin species**! This is the power of mapping a variable to `color` inside `aes()`.

## `aes()` outside vs inside `geom_point()`

```{r}
#| eval: false
# Option A: mapping applies to ALL layers
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point()

# Option B: mapping applies ONLY to this geom layer
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species))
```

Both give the same result here — but Option B is useful once we start **layering multiple geoms** (Session 2!).

## Your Turn 3 (7 min) {.smaller}

Make a scatter plot of `flipper_length_mm` vs `body_mass_g`, colouring the points by `island` instead of `species`.

. . .

**Answer**

```{r}
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, color = island)) +
  geom_point()
```

Notice the colours are **more mixed up** here than with species — island doesn't separate the penguins nearly as cleanly. That itself is a useful finding!

# Bar Charts — Counting Categories

## When to use a bar chart

- Scatter plots need **two numeric variables**
- When you want to **count how many rows fall into each category**, use `geom_bar()`

```{r}
ggplot(data = penguins, aes(x = species)) +
  geom_bar()
```

. . .

Notice: we only supplied **one** aesthetic (`x`). `geom_bar()` automatically counts rows for us — we didn't need to calculate counts ourselves!

## Colouring bars

```{r}
ggplot(data = penguins, aes(x = species, fill = species)) +
  geom_bar()
```

. . .

For bars/areas we usually use **`fill`** (the inside colour) rather than **`color`** (the outline colour). Try swapping `fill` for `color` above and see the difference!

## Your Turn 4 (8 min) {.smaller}

1. Make a bar chart showing how many penguins were recorded on each `island`
2. Fill the bars by `species` — what do you notice about which species live on which islands?

. . .

**Answer**

```{r}
ggplot(data = penguins, aes(x = island, fill = species)) +
  geom_bar()
```

Torgersen only has Adelie penguins; Biscoe has mostly Gentoo (and some Adelie); Dream has Adelie and Chinstrap. Stacked bars like this let us see **two categorical variables at once**.

## Recap — Session 1 {.smaller}

::: {.incremental}
- R is the engine, RStudio is the dashboard
- `<-` stores values in named objects
- `install.packages()` once, `library()` every session
- `penguins` from `palmerpenguins` is our practice dataset all workshop
- Every `ggplot2` plot = **data + aes + geom**, joined with `+`
- `geom_point()` → scatter plots (two numeric variables)
- `geom_bar()` → counts of a categorical variable
- `color`/`fill` inside `aes()` map a variable to colour
:::

## Before Session 2 {.smaller}

**Practice at home (10 min):**

```{r}
#| eval: false
# 1. Make a scatter plot of bill_length_mm vs flipper_length_mm
# 2. Colour it by sex
# 3. Make a bar chart of `sex` counts, coloured by species
```

Bring any error messages you get — we will debug them together in Session 2!

**Next session:** aesthetics deep-dive (size, shape, transparency), histograms, boxplots, density plots, and facets — splitting one plot into a grid of small multiples.

## Thank you! {.center}

**Session 1 complete** 🎉

Questions? Errors on your screen? Raise your hand — we fix them together, not alone.

*SARA Institute of Data Science — building open, first-generation, Ambedkarite research capacity, one dataset at a time.*
