---
title: "Data Visualisation with R"
subtitle: "Session 2 · Grammar of Graphics — Aesthetics, Geoms & Facets"
author: "SARA Institute of Data Science"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    incremental: false
    toc: false
    footer: "SARA Institute of Data Science · R for Researchers"
    scrollable: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(palmerpenguins)
data(penguins)
```

## Welcome back! {.center}

::: {.incremental}
- Session 1 recap: `ggplot(data, aes(x, y)) + geom_...()`
- Today we go deeper into the **grammar**:
  - more aesthetics (size, shape, alpha, transparency)
  - more geoms (histogram, boxplot, density, smooth line)
  - **facets** — many small plots from one line of code
- By the end: you'll be able to explore almost any dataset visually
:::

## Quick recap exercise (5 min)

Recreate this from memory (don't peek at Session 1 notes yet!):

> A scatter plot of `bill_length_mm` (x) vs `bill_depth_mm` (y), coloured by `species`.

. . .

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point()
```

## Session 2 roadmap {.smaller}

| Time | Topic |
|---|---|
| 0–10 min | Recap + more aesthetics (size, shape, alpha) |
| 10–25 min | Trend lines with `geom_smooth()` |
| 25–45 min | Distributions: histograms & density plots |
| 45–65 min | Comparing groups: boxplots |
| 65–85 min | Facets — small multiples |
| 85–90 min | Wrap-up |

# More Aesthetics

## Beyond colour: size, shape, and alpha

Aesthetics aren't limited to `x`, `y`, and `color`. We can also map data to:

- **`size`** — bigger/smaller points
- **`shape`** — different point symbols
- **`alpha`** — transparency (0 = invisible, 1 = solid)

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g,
                     color = species, size = bill_length_mm)) +
  geom_point(alpha = 0.6)
```

. . .

Notice: `alpha = 0.6` is set **outside** `aes()`, directly inside `geom_point()`. That's because we're not mapping it to a *variable* — we're setting **one fixed value** for every point. This is a key distinction in `ggplot2`.

## Mapped vs. fixed — the golden rule {.smaller}

| Inside `aes()` | Outside `aes()` |
|---|---|
| Value **comes from a column** in your data | Value is **the same for every point** |
| `aes(color = species)` | `color = "steelblue"` |
| `aes(size = bill_length_mm)` | `size = 3` |
| Varies per row | Fixed by you |

. . .

```{r}
# Fixed color and size for ALL points - no legend needed
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(color = "darkgreen", size = 2, alpha = 0.7)
```

## Your Turn 1 (7 min) {.smaller}

1. Plot `bill_length_mm` vs `bill_depth_mm`
2. Map `shape` to `species`
3. Set a **fixed** `alpha = 0.5` and `size = 2` (not mapped — just typed directly)

. . .

**Answer**

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, shape = species)) +
  geom_point(alpha = 0.5, size = 2)
```

# Trend Lines

## Adding a trend line with `geom_smooth()`

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  geom_smooth()
```

. . .

Two geoms, **one plot** — this is layering. `geom_smooth()` fits a curve through the data (a "loess" curve by default) with a shaded confidence band.

## A straight-line trend instead

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm")
```

`method = "lm"` fits a straight **l**inear **m**odel instead of a curve — useful when you specifically want to show a linear relationship.

## Trend lines per group

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

. . .

Because `color = species` was set in the shared `aes()`, **both** `geom_point()` and `geom_smooth()` inherit it — giving us one trend line **per species** automatically!

## Your Turn 2 (8 min) {.smaller}

Plot `bill_length_mm` (x) vs `bill_depth_mm` (y):

1. Points coloured by `species`
2. Add a linear (`"lm"`) trend line — one per species

*What do you notice about the overall trend vs. each species' trend? (This is a classic case of Simpson's Paradox!)*

. . .

**Answer**

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Overall, bill length and depth look *negatively* related — but **within each species**, the relationship is *positive*! Always check whether hidden groups are driving a trend.

# Exploring Distributions

## Histograms — one numeric variable

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram()
```

. . .

R gives a friendly warning about bin count — histograms **group numeric values into bins** and count how many rows fall in each.

## Controlling bin width

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, fill = "steelblue", color = "white")
```

. . .

- Smaller `binwidth` → more detail, more noise
- Larger `binwidth` → smoother, but may hide real patterns
- There is no single "correct" binwidth — try a few!

## Histograms by group

```{r}
ggplot(penguins, aes(x = body_mass_g, fill = species)) +
  geom_histogram(binwidth = 250, alpha = 0.6, position = "identity")
```

`position = "identity"` lets the three histograms **overlap** rather than stack — combined with `alpha`, we can see all three at once.

## Density plots — a smooth alternative

```{r}
ggplot(penguins, aes(x = body_mass_g, fill = species)) +
  geom_density(alpha = 0.5)
```

. . .

A density plot is like a **smoothed histogram** — useful for comparing shapes of distributions across groups without worrying about binwidth.

## Your Turn 3 (8 min) {.smaller}

1. Make a histogram of `flipper_length_mm` with `binwidth = 5`
2. Fill it by `species`, with `alpha = 0.6` and `position = "identity"`

. . .

**Answer**

```{r}
ggplot(penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(binwidth = 5, alpha = 0.6, position = "identity")
```

# Comparing Groups

## Boxplots

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

. . .

**Reading a boxplot:**

- box = middle 50% of the data (25th–75th percentile)
- thick line = median
- whiskers = typical range
- dots beyond whiskers = potential outliers

## Boxplots with colour

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.7)
```

Adding `fill = species` is slightly redundant with `x = species` here, but it makes group boundaries easier to see at a glance, especially in printed handouts.

## Adding raw data points on top

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4, color = "darkblue")
```

. . .

`geom_jitter()` nudges overlapping points sideways randomly so we can see individual data — this shows both the **summary** (boxplot) and the **raw data** (jittered points) together.

## Your Turn 4 (8 min) {.smaller}

Make a boxplot of `flipper_length_mm` by `species`, filled by `species`, with jittered points on top.

. . .

**Answer**

```{r}
ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4)
```

# Facets — Small Multiples

## The problem facets solve

- Sometimes colour alone gets **too crowded** with 3+ groups, or we want groups **fully separated**, not overlapping
- **Facets** split one plot into a grid — one panel per category

## `facet_wrap()` — one variable

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  facet_wrap(~species)
```

. . .

The `~species` means "split by species." Each panel shows the **same x/y axes**, just filtered to one species' data.

## `facet_grid()` — two variables at once

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  facet_grid(sex ~ species)
```

. . .

`facet_grid(rows ~ columns)` creates a full grid: rows = `sex`, columns = `species`. Great for comparing **two categorical variables** simultaneously. (Note: a few penguins have missing `sex` — we'll handle missing data more in Session 3.)

## Combining facets with colour

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point() +
  facet_wrap(~island)
```

Facets and aesthetics work together nicely — here we see, **per island**, how species separate by bill measurements.

## Your Turn 5 (10 min) {.smaller}

1. Make a histogram of `body_mass_g`
2. Facet it by `species` using `facet_wrap()`
3. Bonus: also facet by `sex` using `facet_grid(sex ~ species)`

. . .

**Answer**

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, fill = "steelblue") +
  facet_wrap(~species)
```

```{r}
#| eval: false
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, fill = "steelblue") +
  facet_grid(sex ~ species)
```

## Recap — Session 2 {.smaller}

::: {.incremental}
- Aesthetics **mapped** inside `aes()` vary by data; values **set** outside `aes()` are fixed
- `geom_smooth()` adds trend lines — watch for group-level reversals (Simpson's Paradox!)
- `geom_histogram()` / `geom_density()` show the **shape** of one numeric variable
- `geom_boxplot()` compares a numeric variable **across groups**; `geom_jitter()` shows raw data on top
- `facet_wrap(~var)` and `facet_grid(row ~ col)` split one plot into small multiples
:::

## Before Session 3 {.smaller}

**Practice at home:**

```{r}
#| eval: false
# 1. Make a density plot of bill_depth_mm, filled by island
# 2. Facet a boxplot of body_mass_g by species AND sex
# 3. Notice: none of our plots have proper titles yet - we'll fix that next!
```

**Next session:** Titles & labels, themes, custom colour palettes, `coord_flip()`, and — crucially — **saving your plots** with `ggsave()` so they're ready for your paper, poster, or presentation.

## Thank you! {.center}

**Session 2 complete** 🎉

*SARA Institute of Data Science — building open, first-generation, Ambedkarite research capacity, one dataset at a time.*
