---
title: "Data Visualisation with R"
subtitle: "Session 3 · Polishing & Publishing — Labels, Themes, Scales & Saving"
author: "SARA Institute of Data Science"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    incremental: false
    toc: false
    footer: "SARA Institute of Data Science · R for Researchers"
    scrollable: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(palmerpenguins)
data(penguins)
```

## Welcome back! {.center}

::: {.incremental}
- Session 1: grammar basics — `data + aes + geom`
- Session 2: more aesthetics, more geoms, facets
- Session 3 (today): turning a *working* plot into a **publication-ready** plot
  - titles, labels, captions
  - themes and colour palettes
  - flipping coordinates
  - **saving** plots as files for your paper/poster
  - mini project + wrap-up
:::

## Session 3 roadmap {.smaller}

| Time | Topic |
|---|---|
| 0–10 min | Recap + why "finishing touches" matter |
| 10–25 min | Titles, axis labels, captions with `labs()` |
| 25–40 min | Themes — `theme_minimal()`, `theme_bw()`, custom `theme()` |
| 40–55 min | Colour palettes — manual, ColorBrewer, viridis |
| 55–65 min | `coord_flip()` and reordering categories |
| 65–75 min | Saving plots with `ggsave()` |
| 75–90 min | Mini project + wrap-up |

## Why finishing touches matter

- A plot with no title/labels forces the reader to **guess** what they're looking at
- Reviewers, examiners, and funders judge **credibility partly by presentation**
- Good defaults + small, deliberate customisation > over-decorated "chart-junk"

. . .

Today's plots will look noticeably more "professional" — using the *same* grammar you already know, plus a few new functions.

# Titles & Labels

## Adding titles with `labs()`

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Bigger Flippers, Heavier Penguins",
    subtitle = "Relationship between flipper length and body mass",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species",
    caption = "Data: palmerpenguins (Horst, Hill & Gorman)"
  )
```

. . .

`labs()` is another **layer**, added with `+` just like a geom. Every text element a reader sees can be set here: `title`, `subtitle`, `x`, `y`, `color`/`fill` (legend title), and `caption` (source/credit).

## A quick word on data ethics & attribution

- Always **credit your data source** — as we just did in the `caption`
- The `palmerpenguins` data itself credits Dr. Kristen Gorman and the Palmer Station Long Term Ecological Research Network
- Good practice for your own research: cite every dataset you plot, exactly as you'd cite a reference

## Your Turn 1 (8 min) {.smaller}

Take your boxplot of `body_mass_g` by `species` from Session 2, and add:

- a title of your choice
- clear axis labels (not the raw column names!)
- a caption crediting `palmerpenguins`

. . .

**Answer**

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Body Mass Varies Across Penguin Species",
    x = "Species",
    y = "Body mass (g)",
    fill = "Species",
    caption = "Data: palmerpenguins"
  )
```

# Themes

## What is a "theme" in `ggplot2`?

- A **theme** controls the *non-data* ink: background, gridlines, fonts, legend position
- `ggplot2` ships with several ready-made themes

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  theme_minimal()
```

## Comparing built-in themes

```{r}
#| eval: false
theme_gray()     # the default
theme_minimal()  # clean, no border, light gridlines
theme_bw()       # black & white, gridlines kept
theme_classic()  # only x/y axis lines, no gridlines
theme_void()     # nothing but the data - useful for maps/art
```

. . .

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  theme_classic()
```

## Fine-tuning with `theme()`

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 16)
  ) +
  labs(title = "Flipper Length vs Body Mass")
```

. . .

`theme()` is for **fine control** (font size, legend position, etc.), while `theme_minimal()` etc. set a whole **package** of defaults at once. Use a built-in theme first, then tweak individual elements only if needed.

## Your Turn 2 (7 min) {.smaller}

Take any plot from earlier and:

1. Apply `theme_bw()`
2. Move the legend to the `"top"`

. . .

**Answer**

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point() +
  theme_bw() +
  theme(legend.position = "top")
```

# Colour Palettes

## Manual colours

```{r}
ggplot(penguins, aes(x = species, fill = species)) +
  geom_bar() +
  scale_fill_manual(values = c("Adelie" = "#E69F00",
                                "Chinstrap" = "#56B4E9",
                                "Gentoo" = "#009E73"))
```

`scale_fill_manual()` lets you specify **exact colours** — useful for matching institutional branding or a journal's style guide.

## Colour-blind-friendly palettes

- Roughly 1 in 12 men have some form of colour-vision deficiency — always consider this in research figures
- `scale_*_brewer()` and the `viridis` palettes are built to be colour-blind safe and print well in grayscale

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  scale_color_brewer(palette = "Dark2")
```

## The viridis palette

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = bill_length_mm)) +
  geom_point(size = 2) +
  scale_color_viridis_c()
```

. . .

Notice: `scale_color_viridis_c()` (continuous) is for **numeric** variables like `bill_length_mm`; for categorical variables like `species` you'd use `scale_color_viridis_d()` (discrete).

## Your Turn 3 (8 min) {.smaller}

Recreate a scatter plot of `flipper_length_mm` vs `body_mass_g`, coloured by `species`, using:

1. `scale_color_brewer(palette = "Set2")`
2. Then try `scale_color_viridis_d()` — which do you prefer, and why?

. . .

**Answer**

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  scale_color_brewer(palette = "Set2")
```

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  scale_color_viridis_d()
```

# Flipping & Reordering

## `coord_flip()` — horizontal bars

```{r}
ggplot(penguins, aes(x = species, fill = species)) +
  geom_bar() +
  coord_flip() +
  labs(x = NULL, y = "Count of penguins")
```

Horizontal bars are often easier to read when category **names are long** — common in survey/questionnaire research.

## Reordering categories with `fct_reorder()`

```{r}
penguins |>
  filter(!is.na(body_mass_g)) |>
  group_by(species) |>
  summarise(mean_mass = mean(body_mass_g)) |>
  ggplot(aes(x = fct_reorder(species, mean_mass), y = mean_mass)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = "Species", y = "Mean body mass (g)")
```

. . .

`fct_reorder()` (from `forcats`, part of `tidyverse`) sorts categories **by a numeric value** rather than alphabetically — this is one of the single biggest upgrades you can make to a bar chart's readability.

## A gentle note on the pipe `|>`

- `|>` means **"take the thing on the left, and feed it into the function on the right"**
- `penguins |> filter(...) |> group_by(...) |> summarise(...)` reads like a **recipe**, step by step
- This is `dplyr`/tidyverse data wrangling — a natural next skill after visualisation (a great topic for a future SARA session!)

# Saving Your Plots

## `ggsave()` — from screen to file

```{r}
my_plot <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Flipper Length vs Body Mass", x = "Flipper length (mm)",
       y = "Body mass (g)", color = "Species")

ggsave("penguin_plot.png", plot = my_plot, width = 8, height = 5, dpi = 300)
```

. . .

- Save the plot into an **object** first (`my_plot <-`), then pass it to `ggsave()`
- `dpi = 300` is the standard **print-quality** resolution for papers/posters
- Common formats: `.png` (presentations/web), `.pdf`/`.svg` (papers, scalable, crisp for print)

## Choosing width, height & format {.smaller}

| Use case | Format | Tip |
|---|---|---|
| Slides / Word doc | `.png` | `width/height` in inches, `dpi = 300` |
| Journal submission | `.pdf` or `.tiff` | check journal's exact size/DPI requirements |
| Poster | `.pdf` or `.svg` | scalable, won't blur at large sizes |
| Website | `.png` or `.svg` | smaller file sizes matter |

. . .

```{r}
#| eval: false
ggsave("penguin_plot.pdf", plot = my_plot, width = 8, height = 5)
```

## Your Turn 4 (7 min) {.smaller}

1. Build any polished plot of your choice from today
2. Save it as `my_first_plot.png`, sized 7 x 5 inches, at 300 dpi
3. Find the file in your **Files** pane or working folder — open it!

. . .

**Answer**

```{r}
#| eval: false
final_plot <- ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Body Mass by Species", x = "Species", y = "Body mass (g)")

ggsave("my_first_plot.png", plot = final_plot, width = 7, height = 5, dpi = 300)
```

# Mini Project

## Mini project (20 min) {.smaller}

**Build one complete, polished plot from scratch**, using anything from all three sessions. Your plot should include:

::: {.incremental}
1. A sensible geom for your question (point / bar / box / histogram...)
2. At least one meaningful aesthetic mapping (color/fill/shape)
3. A title, axis labels, and a caption crediting the data source
4. A theme (`theme_minimal()`, `theme_bw()`, etc.)
5. A deliberate colour choice (`scale_*_brewer`, `scale_*_viridis`, or manual)
6. Saved to a file with `ggsave()`
:::

*Work in pairs if you like. We'll ask 2–3 volunteers to share their plot and the one question it answers.*

## One example mini project

```{r}
penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = species, y = body_mass_g, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Male Penguins Are Heavier Across All Three Species",
    x = "Species", y = "Body mass (g)", fill = "Sex",
    caption = "Data: palmerpenguins (Horst, Hill & Gorman)"
  ) +
  theme(legend.position = "bottom")
```

## Recap — all three sessions {.smaller}

::: {.incremental}
- **Session 1:** R & RStudio basics, `<-`, packages, `ggplot(data, aes(x,y)) + geom_...()`
- **Session 2:** more aesthetics (size/shape/alpha), `geom_smooth()`, distributions (`geom_histogram`/`geom_density`), `geom_boxplot()`, facets
- **Session 3:** `labs()`, themes, colour palettes, `coord_flip()`, `fct_reorder()`, `ggsave()`
- You now hold the **full grammar**: data → aesthetics → geometry → facets → scales → theme → labels
:::

## Where to go next {.smaller}

- **Book:** *R for Data Science* (Wickham, Çetinkaya-Rundel & Grolemund) — free online at r4ds.hadley.nz
- **Cheatsheet:** RStudio's `ggplot2` cheatsheet (`Help → Cheatsheets` inside RStudio)
- **Practice:** try these same techniques on your **own research data**
- **Community:** R-Ladies, local R user groups, Stack Overflow's `[ggplot2]` tag
- **Next at SARA:** data wrangling with `dplyr`, and/or a Winter School of Statistics session

## Thank you! {.center}

**All three sessions complete** 🎉

You came in having never written a line of R — you're leaving with the ability to explore and publish your own data visualisations.

*SARA Institute of Data Science — building open, first-generation, Ambedkarite research capacity, one dataset at a time.*

**Feedback / questions:** please share with the SARA team — your input shapes the next Summer School!
